Synthesizing Symmetric Lenses

### 2019-03-01 papers/SymmetricLenses/main.tex
remove discussions
spellcheck and other checks

Synthesizing simple symmetric lenses is substantially more challenging than
synthesizing bijective lenses: Since some of the information on each side
can be ``disconnected'' from the other side, there will, in general, be {\em
  many} lenses that agree with a given example. To guide the search process,
we use {\em stochastic regular expressions} and ideas from information
theory to estimate the amount of information propagated by a candidate
lens, generally prefering lenses that propagate more information, 
as well as user annotations marking parts of the source and
target data structures as either {\em irrelevant} or {\em essential}.

We describe an implementation of simple symmetric lenses and our synthesis
procedure as extensions to the Boomerang language. We evaluate its performance
on 48 benchmark examples drawn from Flash Fill, Augeus, the bidirectional
programming literature, and electronic file format synchronization tasks. Our
implementation can synthesize each of these lenses in under 30 seconds.
\end{abstract}

@@

However, the set of real-world use cases for bijective lenses, where two
data formats contain different arrangements of precisely the same
information, is  
limited.  Oftentimes, two data
formats share just {\em some} of their information content.  For instance,
one ad-hoc system-configuration
file might include some format-specific metadata, such as a date or a
reference number, while the same configuration file on another operating
system does not.  Indeed among the 48 benchmark problems described in
Section \ref{sec:evaluation} below, all of the benchmarks taken from Flash
Fill~\cite{flashfill} and many of the benchmarks that synchronize between
two ad hoc file formats have this characteristic.

Can Optician's basic synthesis procedure be extended to a richer class of
lenses? Could we even imagine synthesizing all \emph{symmetric
  lenses}~\cite{symmetric-lenses}---a much larger class that includes
bijective lenses, ``asymmetric'' lenses (where the transformation from a
source structure to a target can throw away information that must
then be restored when transferring from target to source), and even more flexibile
transformations that allow each side to throw away information?

@@

The first of these is pragmatic.  Symmetric lenses as presented
by~\citet{symmetric-lenses} operate
over three structures: a ``left'' structure $X$, a ``right'' structure
$Y$ and a ``complement'' 
$C$ that contains the information not
present in either $X$ or $Y$. These complements must be stored
and managed somehow.  More fundamentially, complements complicate
synthesis {\em specifications}---instead of just giving single examples of
source and target 
pairings, users would have to give longer ``interaction sequences'' to 
show how a lens should behave. To avoid these complexities, we define a
restricted variant 
of symmetric lenses,  called \emph{simple symmetric lenses}. 
%only the usual source and target.
%
%, which we define in Section~\ref{sec:relationship}.
Intuitively, simple symmetric lenses are symmetric
lenses that do not require external ``memory'' to recover data from past
instances of $X$ or $Y$ when making a round trip. They only need the most
recent instance.

@@

The round-tripping laws guarantee that pushing an ``un-edited'' value from one
side (the result of a put or create) back through a lens in the other direction
will get back to exactly where we began. For example, consider a lens
synchronizing the employee data formats in
Figure~\ref{fig:minimized-representations}. Applying the \CreateR function to a
salary file creates a insurance file with the same employee names in the file,
with
\lstinline{"UNK"} for each insurance company. The round-tripping
laws guarantees that applying the \PutL function to the generated insurance file
and the original salary file will return the original salary file, unmodified. If \KW{put}ting a string $s$ into a string $t$ results in the original string $s$, then $s$ and $t$ are considered \emph{synchronized}.

@@

\bcp{I wonder whether the rest of this subsection can be cut... It is both
  hard to understand and too brief to really say much, and I wonder whether
  we need it for anything between here and where we give it a real
  explanation.} \afm{I believe it is necessary for 2.4. I tried giving it a more
terse treatment, but that caused its own set of problems.}

Lenses are typed by pairs of regular expressions and only a handful of
rules result in a given type.  For example, the type of a
\ConcatLens lens is the concatenation of the regular expressions of the sub-lenses.
\[
  \inferrule*
  {
    \Lens_1 \OfType \BRegex_1 \Leftrightarrow \BRegexAlt_1\\
    \Lens_2 \OfType \BRegex_2 \Leftrightarrow \BRegexAlt_2
  }
  {
    \ConcatLensOf{\Lens_1}{\Lens_2} \OfType \BRegex_1 \Concat \BRegex_2
    \Leftrightarrow
    \BRegexAlt_1 \Concat \BRegexAlt_2
  }
\]
The typing  rules for maintain structural similarity
between regular expression types and lens terms, with one exception: The
type equivalence rule permits a well-typed lens to be well-typed among any
equivalent regular expression pair.
%
\[
  \centering
  \inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt\\
    \BRegex \SSREquiv \BRegex'\\
    \BRegexAlt \SSREquiv \BRegexAlt'
  }
  {
    \Lens \OfType \BRegex' \Leftrightarrow \BRegexAlt'
  }
\]
%
This rule is needed for lenses like the final one in
Figure~\ref{fig:example_lens} to be well-typed.

@@

For larger and more complex formats, it can become quite difficult to
write lenses by hand, as we did above.  Instead, we would like to synthesize
them from types and examples.  
That is, given a pair of regular expressions and a set of input-output
examples, we want 
to find a simple symmetric lens that is typed by the regular expression pair and
satisfies all the input/output examples. We this a \emph{satisfying
  lens}. For our running example, suppose we wish to synthesize a lens between
\lstinline{emp_salaries} and \lstinline{emp_insurance}, using as sole
input-output example the data in Figure~\ref{fig:minimized-representations}.

One challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disconnect(emp_salaries, emp_insurance, ..., ...)
\end{lstlisting}
%
is another well-typed lens that satisfies the example in
Figure~\ref{fig:minimized-representations} (where \lstinline{...} represents
the example strings). In general, {\em many} examples
may be 
required to rule out all possible occurrences of \Disconnect lenses,
particularly in complex formats.
%\ksf{I don't understand the following.  We shouldn't explain the issue in
%terms of \PutR and \PutL as we are focused on the combinators rather
%than the raw functions.}
%The default strings given to \Disconnect merely need to be the strings
%in the input-output examples, and the lens will still satisfy the
%specification.  While the disconnect lens is prohibited by two
%examples, or by a carefully chosen input-output example on \PutR
%or \PutL, this problem continues presenting itself for the sublenses,
%and is highlighted even more on larger formats.
Instead of merely finding
\emph{any} satisfying lens, we wish to synthesize a satisfying lens that is
likely to please the user.

@@

The expected number of bits required to recover a piece of data corresponds to
the well-known information-theoretic concept of
\emph{entropy}~\cite{Shannon1948}. Calculating entropy requires a probability
distribution over the space of possible values for the data. Specifically, given
a set $S$ and a probability distribution $P \OfType S \rightarrow \Reals$ over
$S$, the entropy $\EntropyOf{S,P}$ is $-\Sigma_{s\in S}P(s)\Log_2P(s)$. In
information theory, the \emph{information content} of each element of $s \in
S$ is 
the number of bits required to specify $s$ in a perfect encoding scheme
($-\Log_2P(s)$). The entropy of $S$ is the expected number of bits
required to specify an element drawn from $S$---the
probability of each element times its information content. Entropy captures
the intuition that, if a data source contains many possible elements and
none have 
significantly higher probability than others, it will have high information
content. Data sources where most elements have small
probabilities have small entropy. When $P$ is clear from context, we will
often use the shorthand of $\EntropyOf{S}$ for $\EntropyOf{S,P}$.

@@

To obtain SREs from the plain regular expressions that users write, 
we use a default heuristic that attempts to assign probability annotations
giving
each string in the language equal probablity (not prioritizing one piece of
information over another).

Sometimes users already know that certain data should or should not be
used to construct the other format. We introduce relevance annotations to SREs
that enable users to specify whether a piece of data should be used to construct
the other format (with \SRequire) or not (with \Skip). In our example,
\lstinline{salary} and \lstinline{company} could safely be skipped (as they are
disconnected), where \lstinline{name} in \lstinline{emp_salary} and
\lstinline{emp_ins} be annotated as required (as they are converted with the
identity lens). Doing so both constrains the problem (as \lstinline{name}s must
be synchronized) and makes it easier (as the algorithm does not waste time
looking for \lstinline{salary} information in the insurance format). In this
way, users can tweak the
lens likelihoods with their external knowledge.

@@

Following existing work on bidirectional program
synthesis~\cite{maina+:quotient-synthesis}, we 
split our algorithm into two communicating
searche procedures. The first,
\Expand, navigates the space of semantically equivalent regular expressions
by applying rewrite rules that preserve both semantics and
probability distributions. This
algorithm ranks pairs of stochastic regular expressions by the number of
rewrite rule applications required to obtain each pair from the one given as
input. It passes the pairs off to the second 
search procedure, \GreedySynth, in rank order with the smallest first.

\GreedySynth looks for highest-likelihood lenses between a given
pair of stochastic regular expressions $\Regex$ and $\RegexAlt$ by performing a
type-directed search. It first converts the stochastic
regular expressions provided by \Expand into \emph{stochastic DNF regular
  expressions}---a constrained representation of stochastic regular expressions
with disjunctions distributed over concatenations and with concatenations and
disjunctions normalized to operating over lists.
Then it uses the syntax of these n-ary DNF regular expressions to find
normalized lenses in a form we call \emph{simple symmetric n-ary DNF lenses}.
These involve neither a composition operator nor a type equivalence rule.
These restrictions mean that there are only a few simple symmetric n-ary DNF
lenses that are well typed between a given pair of stochastic n-ary DNF regular
expressions, so \GreedySynth's search space for a given pair of regular
expressions is finite. Finally, \GreedySynth yields a simple symmetric lens by
converting the n-ary syntax back to the binary forms provided in the surface
language.

This architecture of communicating synthesizers gives us a way to enumerate
pairs of stochastic regular expressions of increasing rank and to efficiently
search through them, but it poses a problem: when should \Expand stop proposing
new SRE pairs? We might have found a promising lens between a pair of stochastic
regular expressions, but a different pair we haven't yet discovered may give
rise to an even better lens. The search algorithm must resolve a tension between
the quality of the inferred lens and the amount of time it takes to return a
result. For example, if the algorithm has already found the lens in
Figure~\ref{fig:example_lens}, we don't want to spend a lot of time searching
for an even better lens. To resolve this tension, the algorithm uses heuristics
to judge whether to return the current best satisfying lens to the user or to
pass the next set of unrollings to \GreedySynth. The heuristics favor stopping
if the current best satisfying lens is very likely, indicating the lens is very
promising (for example, if the satisfying lens loses no information, the
algorithm should terminate for no other lens will be more likely). The
heuristics also favor stopping if \Expand has delivered to \GreedySynth all
pairs of stochastic regular expressions at a given rank and there is a large
number of pairs at the next rank, because searching through all such pairs will
take a long time (for example, if the satisfying lens loses only a little
information,
but searching for a better one will take a long time, the discovered lens is
returned).
With this approach, the algorithm can quickly return a
satisfying lens with relatively high likelihood. If the user is unhappy with the
result, they can either refine the search by supplying additional examples,
which serve both to rule out previously proposed lenses and to reduce the
size of the search space by cutting down on the number of satisfying lenses, or
they can supply annotations on the source and target SREs indicating that
certain information either must be retained or must be discarded by the lens
(see \S\ref{subsec:relevanceannotations}).

@@

For example, the entropy of $\Regex =
\PRegexOr{\lstinline{"a"}}{(\lstinline{"b"})}{.5}$ is 1. The best encoding of a
stream of elements from $\Regex$ will use, on average, 1 bit per element to
determine whether that element is \lstinline{"a"} or \lstinline{"b"}. As an
additional example, a fixed string has no information content, and so has no
entropy.

@@

    $\REntropyOf{\Regex \Given \InvertOf{\Lens}, \RegexAlt}$
    & =
    & $\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$\\
  \end{tabular}
\end{center}
$\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$ is defined symmetrically. These
functions bound the expected number of bits to recover one data format from a
synchronized string in the other format. Note that we would be able to exactly
calculate the conditional entropy, were it not for \MergeL and \MergeR. If
$\MergeLOf{\Lens_1}{\Lens_2} \OfType \Regex \Leftrightarrow
\PRegexOr{\RegexAlt_1}{\RegexAlt_2}{\ProbabilityAlt}$, given a string in $s$, we
need to determine if the synchronized string is in $\RegexAlt_1$ or $\RegexAlt_2$.
However, this information content is dependent on the how likely the
synchronized string is to be in $\RegexAlt_1$ or $\RegexAlt_2$. Nevertheless, we
typically calculate the conditional entropy exactly, as merges are relatively
uncommon in practice; only 2 of the lenses sythesized in our benchmarks
include merges.

@@

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
Algorithm~\ref{alg:synth-sym-lens} presents our synthesis algorithm at a high
level of abstraction. This algorithm searches for likely lenses in priority
order one ``class'' at a time using a \GreedySynth subroutine. Each class is the
set of lenses that can by typed by a pair of regular expressions, modulo a set
of simple axioms such as associativity, commutativity, and distributivity. The
\Expand subroutine generates new classes using the star-unrolling axioms
(\UnrollstarLeftRule and \UnrollstarRightRule).

To summarize, the input regular
expressions are first converted into stochastic regular expressions with
\ToStochastic. This pair of SREs is used to initialize a priority queue ($pq$).
The priority of a SRE pair is the number of rewrites needed to derive the pair
from the originals. Next, $\PCF{SynthSymLens}$ enters a loop that searches for
likely lenses. The loop terminates when the algorithm believes it is unlikely to
find a better lens than the best one it has found so far (a termination
condition defined by $\PCF{\Continue}$). Within each iteration of the loop, it:
\begin{itemize}
\item pops the next class ($S$, $T$) of lenses to
search off of the priority queue ($\PCF{\PQ.Pop}$),
\item executes $\GreedySynth$ to find a best lens in that class if
one exists ($\Lens$),
using the examples $exs$ to filter out potential lenses that do not satisfy
the specification,
\item replaces $best$ with $\Lens$, if $\Lens$ is more likely according
to our information-theoretic metric, and
\item adds the SREs derived from rewriting $S$ and $T$ ($\Expand(S,T)$)
  to the priority queue.
\end{itemize}
When the loop terminates, the search returns the globally best lens found
($best$).  Each subroutine of this algorithm will be explained in
further depth in the following subsections.

@@

Similarly, users can recognize that employee names are present in both files. By
annotating instances of \lstinline{name} as \lstinline{require(name)}, users can
add this knowledge to the specification to optimize the search and force the
the generated lens to retain \lstinline{name} information. In practice, we
make any lens that loses ``required'' information infinitely unlikely.
%
\begin{center}
  \begin{tabular}{rcl}
    $\REntropyOf{\SRequireOf{\RegexAlt} \Given \Lens, \Regex}$
    & =
    & $\begin{cases}
      \infty & \text{if }\REntropyOf{\SRequireOf{\RegexAlt} \Given \Lens, \Regex}
      \neq 0\\
    0 & \text{otherwise}
  \end{cases}$\\

@@

\subsection{\GreedySynth}
\label{subsec:greedy-synth}
The synthesis procedure comprises three algorithms: one that greedily finds
symmetric DNF lenses (\GreedySynth), one that greedily finds symmetric sequence
lenses (\GreedySeqSynth), and one that finds symmetric atom lenses
(\AtomSynth{}). These three algorithms are hierarchically structured:
\GreedySynth relies on \GreedySeqSynth, \GreedySeqSynth relies on \AtomSynth,
and \AtomSynth rlies on \GreedySynth. The structure of the algorithms mirrors
the structure of symmetric DNF lenses and \SDNFREabbrevs.

\begin{algorithm}[t]
  \caption{\GreedySynth}
  \label{alg:greedysynth}
  \begin{algorithmic}[1]
    \Function{GreedySynth}{$\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}$}
    \If{$\Call{CannotMap}{\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}}$}
    \State $\ReturnVal{\None}$
    \EndIf
    \State $\SLS \gets
    \Call{\CartesianMap}{\GreedySeqSynth(\Examples),\ListOf{\Sequence_1;\ldots;\Sequence_n},\ListOf{\SequenceAlt_1;\ldots;\SequenceAlt_m}}$
    \State $\GreedyState \gets \Call{PQ.Create}{\SLS}$
    \State $lb \gets \Call{LensBuilder.Empty}{}$
    \While{$\Call{PQ.IsNonempty}{\GreedyState}$}
    \State $\SSQLens \gets
    \Call{PQ.Pop}{\GreedyState}$
    \If{$\Call{LensBuilder.UsefulAdd}{lb,\SSQLens,\Examples}$}
    \State $lb \gets \Call{LensBuilder.AddSeq}{lb,\SSQLens}$
    \EndIf
    \EndWhile
    \State \ReturnVal{\Call{LensBuilder.ToDNFLens}{\GreedyState}}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

>>>>

Synthesizing simple symmetric lenses is substantially more challenging than
synthesizing bijective lenses: Since some of the information on each side
can be ``disconnected'' from the other side, there will, in general, be {\em
  many} lenses that agree with a given example. To guide the search process,
we use {\em stochastic regular expressions} and ideas from information
theory to estimate the amount of information propagated by a candidate
lens, generally preferring lenses that propagate more information, 
as well as user annotations marking parts of the source and
target data structures as either {\em irrelevant} or {\em essential}.

We describe an implementation of simple symmetric lenses and our synthesis
procedure as extensions to the Boomerang language. We evaluate its performance
on 48 benchmark examples drawn from Flash Fill, Augeas, the bidirectional
programming literature, and electronic file format synchronization tasks. Our
implementation can synthesize each of these lenses in under 30 seconds.
\end{abstract}

@@

However, the set of real-world use cases for bijective lenses, where two
data formats contain different arrangements of precisely the same
information, is  
limited.  Oftentimes, two data
formats share just {\em some} of their information content.  For instance,
one ad-hoc system-configuration
file might include some format-specific metadata, such as a date or a
reference number, while the same configuration file on another operating
system does not.  Indeed among the benchmark problems described in
Section \ref{sec:evaluation}, all of the benchmarks taken from Flash
Fill~\cite{flashfill} and many of the benchmarks that synchronize between
two ad hoc file formats have this characteristic.

Can Optician's basic synthesis procedure be extended to a richer class of
lenses? Could we even imagine synthesizing all \emph{symmetric
  lenses}~\cite{symmetric-lenses}---a much larger class that includes
bijective lenses, ``asymmetric'' lenses (where the transformation from a
source structure to a target can throw away information that must
then be restored when transferring from target to source), and even more flexible
transformations that allow each side to throw away information?

@@

The first of these is pragmatic.  Symmetric lenses as presented
by~\citet{symmetric-lenses} operate
over three structures: a ``left'' structure $X$, a ``right'' structure
$Y$ and a ``complement'' 
$C$ that contains the information not
present in either $X$ or $Y$. These complements must be stored
and managed somehow.  More fundamentally, complements complicate
synthesis {\em specifications}---instead of just giving single examples of
source and target 
pairings, users would have to give longer ``interaction sequences'' to 
show how a lens should behave. To avoid these complexities, we define a
restricted variant 
of symmetric lenses,  called \emph{simple symmetric lenses}. 
%only the usual source and target.
%
%, which we define in Section~\ref{sec:relationship}.
Intuitively, simple symmetric lenses are symmetric
lenses that do not require external ``memory'' to recover data from past
instances of $X$ or $Y$ when making a round trip. They only need the most
recent instance.

@@

The round-tripping laws guarantee that pushing an ``unedited'' value from one
side (the result of a put or create) back through a lens in the other direction
will get back to exactly where we began. For example, consider a lens
synchronizing the employee data formats in
Figure~\ref{fig:minimized-representations}. Applying the \CreateR function to a
salary file creates a insurance file with the same employee names in the file,
with
\lstinline{"UNK"} for each insurance company. The round-tripping
laws guarantees that applying the \PutL function to the generated insurance file
and the original salary file will return the original salary file, unmodified.

@@

Lenses are typed by pairs of regular expressions and only a handful of
rules result in a given type.
Many of these rules provide tight relationships between terms and types, for example, the type of a
\ConcatLens lens is the concatenation of the regular expressions of the sub-lenses.
\[
  \inferrule*
  {
    \Lens_1 \OfType \BRegex_1 \Leftrightarrow \BRegexAlt_1\\
    \Lens_2 \OfType \BRegex_2 \Leftrightarrow \BRegexAlt_2
  }
  {
    \ConcatLensOf{\Lens_1}{\Lens_2} \OfType \BRegex_1 \Concat \BRegex_2
    \Leftrightarrow
    \BRegexAlt_1 \Concat \BRegexAlt_2
  }
\]
The typing  rules for maintain structural similarity
between regular expression types and lens terms, with one exception: The
type equivalence rule permits a well-typed lens to be well-typed among any
equivalent regular expression pair.
%
\[
  \centering
  \inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt\\
    \BRegex \SSREquiv \BRegex'\\
    \BRegexAlt \SSREquiv \BRegexAlt'
  }
  {
    \Lens \OfType \BRegex' \Leftrightarrow \BRegexAlt'
  }
\]
%
This rule is needed for lenses like the final one in
Figure~\ref{fig:example_lens} to be well-typed.

@@

For larger and more complex formats, it can become quite difficult to
write lenses by hand, as we did above.  Instead, we would like to synthesize
them from types and examples.  
That is, given a pair of regular expressions and a set of input-output
examples, we want 
to find a simple symmetric lens that is typed by the regular expression pair and
satisfies all the input/output examples. We call this a \emph{satisfying
  lens}. For our running example, suppose we wish to synthesize a lens between
\lstinline{emp_salaries} and \lstinline{emp_insurance}, using as sole
input-output example the data in Figure~\ref{fig:minimized-representations}.

One challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disconnect(emp_salaries, emp_insurance, $s$, $t$)
\end{lstlisting}
%
is another well-typed lens that satisfies the example in
Figure~\ref{fig:minimized-representations} (where $s$ and $t$ represent
the provided example strings). In general, {\em many} examples
may be 
required to rule out all possible occurrences of \Disconnect lenses,
particularly in complex formats.
%\ksf{I don't understand the following.  We shouldn't explain the issue in
%terms of \PutR and \PutL as we are focused on the combinators rather
%than the raw functions.}
%The default strings given to \Disconnect merely need to be the strings
%in the input-output examples, and the lens will still satisfy the
%specification.  While the disconnect lens is prohibited by two
%examples, or by a carefully chosen input-output example on \PutR
%or \PutL, this problem continues presenting itself for the sublenses,
%and is highlighted even more on larger formats.
Instead of merely finding
\emph{any} satisfying lens, we wish to synthesize a satisfying lens that is
likely to please the user.

@@

The expected number of bits required to recover a piece of data corresponds to
the well-known information-theoretic concept of
\emph{entropy}~\cite{Shannon1948}. Calculating entropy requires a probability
distribution over the space of possible values for the data. Specifically, given
a set $S$ and a probability distribution $P \OfType S \rightarrow \Reals$ over
$S$, the entropy $\EntropyOf{S,P}$ is $-\Sigma_{s\in S}P(s)\Log_2P(s)$. In
information theory, the \emph{information content} of each element of $s \in
S$ is 
the number of bits required to specify $s$ in a perfect encoding scheme
($-\Log_2P(s)$). The entropy of $S$ is the expected number of bits
required to specify an element drawn from $S$---the
probability of each element times its information content. Entropy captures
the intuition that, if a data source contains many possible elements and
none have 
significantly higher probability than others, it will have high entropy. Data
sources with just a few high probability elements have low
entropy. When $P$ is clear from context, we will
often use the shorthand of $\EntropyOf{S}$ for $\EntropyOf{S,P}$.

@@

To obtain SREs from the plain regular expressions that users write, 
we use a default heuristic that attempts to assign probability annotations
giving
each string in the language equal probability (not prioritizing one piece of
information over another).

Sometimes users already know that certain data should or should not be
used to construct the other format. We introduce relevance annotations to SREs
that enable users to specify whether a piece of data should be used to construct
the other format (with \SRequire) or not (with \Skip). In our example,
\lstinline{salary} and \lstinline{company} could safely be skipped (as they are
disconnected), where \lstinline{name} in \lstinline{emp_salary} and
\lstinline{emp_ins} could be annotated as required (as they are converted with the
identity lens). Doing so both constrains the problem (as \lstinline{name}s must
be synchronized) and makes it easier (as the algorithm does not waste time
looking for \lstinline{salary} information in the insurance format). In this
way, users can tweak the
lens likelihoods with their external knowledge.

@@

Following existing work on bidirectional program
synthesis~\cite{maina+:quotient-synthesis}, we 
split our algorithm into two communicating
search procedures. The first,
\Expand, navigates the space of semantically equivalent regular expressions
by applying rewrite rules that preserve both semantics and
probability distributions. This
algorithm ranks pairs of stochastic regular expressions by the number of
rewrite rule applications required to obtain each pair from the one given as
input. It passes the pairs off to the second 
search procedure, \GreedySynth, in rank order with the smallest first.

\GreedySynth looks for highest-likelihood lenses between a given
pair of stochastic regular expressions $\Regex$ and $\RegexAlt$ by performing a
type-directed search. It first converts the stochastic
regular expressions provided by \Expand into \emph{stochastic DNF regular
  expressions}---a constrained representation of stochastic regular expressions
with disjunctions distributed over concatenations and with concatenations and
disjunctions normalized to operating over lists.
Then it uses the syntax of these n-ary DNF regular expressions to find
normalized lenses in a form we call \emph{simple symmetric n-ary DNF lenses}.
These involve neither a composition operator nor a type equivalence rule.
These restrictions mean that there are comparatively few simple symmetric n-ary DNF
lenses that are well typed between a given pair of stochastic n-ary DNF regular
expressions, so \GreedySynth's search space for a given pair of regular
expressions is finite. Finally, \GreedySynth yields a simple symmetric lens by
converting the n-ary syntax back to the binary forms provided in the surface
language.

This architecture of communicating synthesizers gives us a way to enumerate
pairs of stochastic regular expressions of increasing rank and to efficiently
search through them, but it poses a problem: when should \Expand stop proposing
new SRE pairs? We might have found a promising lens between a pair of stochastic
regular expressions, but a different pair we haven't yet discovered may give
rise to an even better lens. The search algorithm must resolve a tension between
the quality of the inferred lens and the amount of time it takes to return a
result. For example, if the algorithm has already found the lens in
Figure~\ref{fig:example_lens}, we don't want to spend a lot of time searching
for an even better lens. To resolve this tension, the algorithm uses heuristics
to judge whether to return the current best satisfying lens to the user or to
pass the next set of equivalent SREs to \GreedySynth. The heuristics favor stopping
if the current best satisfying lens is very likely, indicating the lens is very
promising (for example, if the satisfying lens loses no information, the
algorithm should terminate for no other lens will be more likely). The
heuristics also favor stopping if \Expand has delivered to \GreedySynth all
pairs of stochastic regular expressions at a given rank and there is a large
number of pairs at the next rank, because searching through all such pairs will
take a long time (\IE, if a satisfying lens loses only a little
information
and searching for a better one will take a long time, the discovered lens is
returned).
With this approach, the algorithm can quickly return a
satisfying lens with relatively high likelihood. If the user is unhappy with the
result, they can either refine the search by supplying additional examples,
which serve both to rule out previously proposed lenses and to reduce the
size of the search space by cutting down on the number of satisfying lenses, or
they can supply annotations on the source and target SREs indicating that
certain information either must be retained or must be discarded by the lens
(see \S\ref{subsec:relevanceannotations}).

@@

For example, the entropy of $\Regex =
\PRegexOr{\lstinline{"a"}}{\lstinline{"b"}}{.5}$ is 1. The best encoding of a
stream of elements from $\Regex$ will use, on average, 1 bit per element to
determine whether that element is \lstinline{"a"} or \lstinline{"b"}. As an
additional example, a fixed string has no information content, and so has no
entropy.

@@

    $\REntropyOf{\Regex \Given \InvertOf{\Lens}, \RegexAlt}$
    & =
    & $\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$\\
  \end{tabular}
\end{center}
$\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$ is defined symmetrically. These
functions bound the expected number of bits to recover one data format from a
synchronized string in the other format. Note that we would be able to exactly
calculate the conditional entropy, were it not for \MergeL and \MergeR. If
$\MergeLOf{\Lens_1}{\Lens_2} \OfType \Regex \Leftrightarrow
\PRegexOr{\RegexAlt_1}{\RegexAlt_2}{\ProbabilityAlt}$, given a string in $s$, we
need to determine if the synchronized string is in $\RegexAlt_1$ or $\RegexAlt_2$.
However, this information content is dependent on the how likely the
synchronized string is to be in $\RegexAlt_1$ or $\RegexAlt_2$. Nevertheless, we
typically calculate the conditional entropy exactly, as merges are relatively
uncommon in practice; only 2 of the lenses synthesized in our benchmarks
include merges.

@@

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
Algorithm~\ref{alg:synth-sym-lens} presents our synthesis algorithm at a high
level of abstraction. This algorithm searches for likely lenses in priority
order one ``class'' at a time using a \GreedySynth subroutine. Each class is the
set of lenses that can by typed by a pair of regular expressions, modulo a set
of simple axioms such as associativity, commutativity, and distributivity. The
\Expand subroutine generates new classes using the star-unrolling axioms
(\UnrollstarLeftRule{} and \UnrollstarRightRule).

To summarize, the input regular
expressions are first converted into stochastic regular expressions with
\ToStochastic. This pair of SREs is used to initialize a priority queue ($pq$).
The priority of a SRE pair is the number of rewrites needed to derive the pair
from the originals. Next, $\PCF{SynthSymLens}$ enters a loop that searches for
likely lenses. The loop terminates when the algorithm believes it is unlikely to
find a better lens than the best one it has found so far (a termination
condition defined by $\PCF{\Continue}$). Within each iteration of the loop, it:
\begin{itemize}
\item pops the next class ($S$, $T$) of lenses to
search off of the priority queue ($\PCF{\PQ.Pop}$),
\item executes $\GreedySynth$ to find a best lens in that class if
one exists ($\Lens$),
using the examples $\Examples$ to filter out potential lenses that do not satisfy
the specification,
\item replaces $best$ with $\Lens$, if $\Lens$ is more likely according
to our information-theoretic metric, and
\item adds the SREs derived from rewriting $S$ and $T$ ($\Expand(S,T)$)
  to the priority queue.
\end{itemize}
When the loop terminates, the search returns the globally best lens found
($best$).  Each subroutine of this algorithm will be explained in
further depth in the following subsections.

@@

Similarly, users can recognize that employee names are present in both files. By
annotating instances of \lstinline{name} as \lstinline{require(name)}, users can
add this knowledge to the specification to optimize the search and force the
generated lens to retain \lstinline{name} information. In practice, we
make any lens that loses ``required'' information infinitely unlikely.
%
\begin{center}
  \begin{tabular}{rcl}
    $\REntropyOf{\SRequireOf{\RegexAlt} \Given \Lens, \Regex}$
    & =
    & $\begin{cases}
      \infty & \text{if }\REntropyOf{\SRequireOf{\RegexAlt} \Given \Lens, \Regex}
      \neq 0\\
    0 & \text{otherwise}
  \end{cases}$\\

@@

\subsection{\GreedySynth}
\label{subsec:greedy-synth}
The synthesis procedure comprises three algorithms: one that greedily finds
symmetric DNF lenses (\GreedySynth), one that greedily finds symmetric sequence
lenses (\GreedySeqSynth), and one that finds symmetric atom lenses
(\AtomSynth{}). These three algorithms are hierarchically structured:
\GreedySynth relies on \GreedySeqSynth, \GreedySeqSynth relies on \AtomSynth,
and \AtomSynth relies on \GreedySynth. The structure of the algorithms mirrors
the structure of symmetric DNF lenses and \SDNFREabbrevs.

\begin{algorithm}[t]
  \caption{\GreedySynth}
  \label{alg:greedysynth}
  \begin{algorithmic}[1]
    \Function{GreedySynth}{$\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}$}
    \If{$\Call{CannotMap}{\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}}$}
    \State $\ReturnVal{\None}$
    \EndIf
    \State $\SLS \gets
    \Call{\CartesianMap}{\GreedySeqSynth(\Examples),\ListOf{\Sequence_1;\ldots;\Sequence_n},\ListOf{\SequenceAlt_1;\ldots;\SequenceAlt_m}}$
    \State $\GreedyState \gets \Call{PQ.Create}{\SLS}$
    \State $\LensBuilder \gets \Call{LensBuilder.Empty}{}$
    \While{$\Call{PQ.IsNonempty}{\GreedyState}$}
    \State $\SSQLens \gets
    \Call{PQ.Pop}{\GreedyState}$
    \If{$\Call{LensBuilder.UsefulAdd}{\LensBuilder,\SSQLens,\Examples}$}
    \State $\LensBuilder \gets \Call{LensBuilder.AddSeq}{\LensBuilder,\SSQLens}$
    \EndIf
    \EndWhile
    \State \ReturnVal{\Call{LensBuilder.ToDNFLens}{\GreedyState}}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\paragraph*{Symmetric DNF Lenses} Algorithm~\ref{alg:greedysynth} presents
\GreedySynth, which synthesizes symmetric DNF lenses. Its inputs 
are a suite of input-output examples and a pair of stochastic DNF regular expressions.
%
First, \PCF{CannotMap} determines whether there is no lens satisfying the
examples, and if so, \GreedySynth returns \None immediately. Otherwise, \GreedySynth finds
the best lenses (given the examples that match them) between all sequence pairs
$(SQ_i, TQ_j)$ drawn from the left and right DNF regular expressions. (The
function \CartesianMap maps its argument across the cross product of the input
lists). A priority queue containing these sequence lenses, ordered by likelihood, is
then initialized with $\PCF{PQ.Create}$. The symmetric lens is then built up
iteratively from these sequence lenses, where the state of the partially
constructed lens is tracked in the lens builder, \LensBuilder.
